Staged Abstract Interpreters

### 2019-02-12 paper/10_intro.tex
code & revision

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections, which first identifies the binding-time of variables in
the program, i.e., they can be known whether statically or dynamically, and then
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-time is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP
for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the stage
annotations (i.e., binding-time annotations) to be explict from the programmers.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we consider the power function:

@@

If the programmer identifies that @b@ will be known in the future stage (as shown
in its type @Rep[Int]@ -- a representation of @Int@) and @e@ is known in the
current stage, say 5, then we can specialize the function @power@ with
respect to @e = 5@ and generate a specialized function that runs faster:

@@

Roughly at the same time when Futamura projections was visioned in 1970s,
\citeauthor{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fixpoints \cite{DBLP:conf/popl/CousotC77}.
Some recent progress such as Abstracting Abstract Machines (AAM)
uncovers a methodology to derive sound abstract interpreters from their concrete
counterparts and has been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}. Given the structural
similarity between concrete and abstract interpreters, intellectually it is
natural to raise the question whether it is possible to specialize an abstract
interpreter, and how can we effectively do that having considered their
different functionalities. This paper studies the confluence of two old ideas
--- Futamura projection and abstract interpretation, but from the perspective of
their recent realizations --- multi-stage programming and definitional abstract
interpreters. To be specific, we present the application of the first Futamura
projection on abstract interpreters, and the approach of constructing optimizing
abstract interpreters by multi-stage programming.

@@

Multi-stage programming has been widely used to improve the performance in many
domains, such as optimizing compilers or domain-specific languages
\cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17, Ofenbeck:2017:SGP:3136040.3136060}, data
processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18}, etc.
Likewise, specializing static analyses by partial evaluation emerged in late 90s
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical}, and indeed it is able to effectively remove the interpretive
overhead of repeatedly traversing the abstract syntax tree. However, these
previous works focused mostly on one particular analysis, or required to
completely rewrite the analyzer. Hence, it is worth to investigate the idea from
a modern perspective based on generative programming, especially for a general
abstract interpreter that models direct-style $\lambda$-calculus, imperative
features such as mutation, and different abstract domains. One technical
challenge is the binding-time engineering when non-determinism, fixed-point
iterations, different abstract domains and staging are introduced at the same
time. In this paper, we present an end-to-end design and implementation of
staging an abstract interpreter; that means not only the interpreter that
traverses the abstract syntax tree, but also the data structure of abstract
domains, abstract environment and heap, and the fixed-point iteration are
staged.

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections, which first identifies the binding-time of variables in
the program, i.e., they can be known whether statically or dynamically, and then
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-time is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP
for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the stage
annotations (i.e., binding-time annotations) to be explicit from the programmers.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we consider the power function to introduce
the idea of MSP:

@@

If the programmer identifies that @b@ will be known in the future stage (as shown
on its type @Rep[Int]@ -- a representation of @Int@) and @e@ is known in the
current stage, say 5, then we can specialize the function @power@ with
respect to @e = 5@ and generate a specialized function that runs faster:

@@

Roughly at the same time when Futamura projections was visioned in 1970s,
\citeauthor{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fixpoints \cite{DBLP:conf/popl/CousotC77}.
Recent progress such as Abstracting Abstract Machines (AAM) uncovers a
methodology to derive sound abstract interpreters from their concrete
counterparts, and it has been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.
Given the structural similarity between concrete and abstract interpreters,
intellectually it is natural to raise the question whether it is possible to
specialize an abstract interpreter, and how can we effectively do that having
considered their different functionalities.

This paper studies the confluence of two old ideas --- Futamura projection and
abstract interpretation, but from the perspective of their recent realizations
--- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract interpreters.
By borrowing the ideas from \note{monadic interpreters} and embedded
domain-specific languages, we develop an approach to fulfill four different
semantics (Figure~\ref{confluence}) in a unified framework.

% FIXME: and the approach of constructing optimizing abstract interpreters by multi-stage programming.

@@

\note{besides the intellectual merit of the confluence diagram, practical
problem: can we speed up a static analyzer without pervasive changes of the
abstract semantics, even without changing a single line of the abstract
interpreter?}

\todo{Move some into related work}
Multi-stage programming has been widely used to improve the performance in many
domains, such as optimizing compilers or domain-specific languages
\cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17, Ofenbeck:2017:SGP:3136040.3136060}, data
processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18}, etc.
Likewise, specializing static analyses by partial evaluation emerged in late 90s
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical}, and indeed it is able to effectively remove the interpretive
overhead of repeatedly traversing the abstract syntax tree. However, these
previous works focused mostly on one particular analysis, or required to
completely rewrite the analyzer. Hence, it is worth to investigate the idea from
a modern perspective based on generative programming, especially for a general
abstract interpreter that models direct-style $\lambda$-calculus, imperative
features such as mutation, and different abstract domains. One technical
challenge is the binding-time engineering when non-determinism, fixed-point
iterations, different abstract domains and staging are introduced at the same
time. In this paper, we present an end-to-end design and implementation of
staging an abstract interpreter; that means not only the interpreter that
traverses the abstract syntax tree, but also the data structure of abstract
domains, abstract environment and heap, and the fixed-point iteration are
staged.

### 2019-02-12 paper/21_bg_interp.tex
code & revision

\begin{lstlisting}
abstract class Expr
case class Lit(i: Int)
case class Var(x: String)
case class Lam(x: String, e: Expr)
case class App(e1: Expr, e2: Expr)
case class AOp(op: String, e1: Expr, e2: Expr)
case class Rec(x: String, e: Expr, body: Expr)
case class If0(cnd: Expr, thn: Expr, els: Expr)
\end{lstlisting}

>>>>

\begin{lstlisting}
abstract class Expr
case class Lit(i: Int)                         // numbers
case class Var(x: String)                      // variables
case class Lam(x: String, e: Expr)             // lambda abstractions
case class App(e1: Expr, e2: Expr)             // applications
case class AOp(op: String, e1: Expr, e2: Expr) // primitive operations
case class Rec(x: String, rhs: Expr, e: Expr)  // recursive bindings
case class If0(e1: Expr, e2: Expr, e3: Expr)   // conditionals
\end{lstlisting}
